---
title: "TINMAN 16S Analysis of Differentially Abundant Taxa (SILVA Data)"
author: "Rachel Rodgers"
date: "`r Sys.Date()`"
output: 
  html_document:
    code_folding: hide
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.width=8,
                      fig.height=6,
                      warning = FALSE,
                      message = FALSE)

knitr::opts_knit$set(root.dir = "..")
```

```{r load-libraries, message=FALSE}
library("here")
library("shiny")
library("knitr")
library("phyloseq")
library("gridExtra")
library("ggrepel")
library("ggpubr")
library("scales")
library("plotly")
library("DESeq2")
library("FSA")
library("writexl")
library("boot")
library("patchwork")
library("tidyverse")

figPath <- "../analysis/figures/16S/"
resultsPath <- "../analysis/results/16S"

source(here("shared_R_scripts/Helper_Functions.R"))
source(here("shared_R_scripts/BoxPlotHelpers.R"))
source(here("shared_R_scripts/BiomarkerHelpers.R"))

options(shiny.sanitize.errors = FALSE)
```

```{r read-in-data, results='hide'}
physeqBacteria <- readRDS(here("../data/RDataObjects/16S/physeqObjects/physeqBacteria.RDS"))
physeqBacteria # 1809 x 76

# modify physeqBacteria sample data slot
sampleDataBacteria <- as(sample_data(physeqBacteria), "data.frame")
sampleDataBacteria$timepoint_condition <- factor(sampleDataBacteria$timepoint_condition,
                                                 levels = c("PRE_Control",
                                                            "POST_Control",
                                                            "PRE_Neutropenic",
                                                            "POST_Neutropenic"))
sample_data(physeqBacteria) <- sample_data(sampleDataBacteria)

taxTableFull <- as.data.frame(as(tax_table(physeqBacteria), "matrix")) %>% 
  rownames_to_column("ASV")

# taxa more abundant in Neutropenic Pre compared to Neutropenic Post
neutropenicTaxaDF <- read.delim(here("../documents/tinman16S_bacterialTaxaOfInterest_neutropenicPrePost.txt"),
                                header = FALSE, col.names = "ASV") %>% 
  merge(., taxTableFull, by = "ASV", all = FALSE)
neutropenicTaxaVec <- pull(neutropenicTaxaDF, ASV)

# taxa more abundant in Control Post compared to Neutropenic Post
postTaxaDF <- read.delim(here("../documents/tinman16S_bacterialTaxaOfInterest_controlNeutropenicPost.txt"),
                       header = FALSE, col.names = "ASV") %>%
  merge(., taxTableFull, by = "ASV", all = FALSE)
postTaxaVec <- pull(postTaxaDF, ASV)

# all taxa combined
allTaxaDF <- rbind(neutropenicTaxaDF, postTaxaDF)
allTaxaVec <- pull(allTaxaDF, ASV)
```

```{r save-allTaxaDF, eval=FALSE, include=FALSE}
saveRDS(allTaxaDF, file = here("../data/RDataObjects/16S/allTaxaDF.RDS"))
```

## DESeq Ground Truth Plots

Ground truth plots are used to visualize the abundance of taxa of interest in
each sample, with samples grouped by variables of interest. This visualization
helps to clarify why DESeq2 may have identified these taxa - are they highly
abundant in one or a few samples (potentially biologically less relevant), or is 
there an  indication of a trend that may be correlated with a biologically 
relevant factor?

```{r PlotGroundTruth}

PlotGroundTruth <- function(abundanceDF, varName, taxaName, subtitleStr) {
  
  plot <- ggplot(abundanceDF,
         aes_string(x = varName, y = "Abundance")) +
    geom_boxplot(outlier.shape = NA) +
    geom_jitter(alpha = 0.6, size = 3.5, width = 0.2) +
    scale_y_log10() +
    ylab("Abundance") +
    xlab(varName) +
    ggtitle(paste("Abundance of\n", taxaName),
            subtitle = subtitleStr) +
    theme_pubr() +
    theme(legend.position = "none",
          axis.text.x = element_text(angle = 45, hjust = 1),
          plot.title = element_text(hjust = 0.5),
          plot.subtitle = element_text(hjust = 0.5))
}
```

### Generate Taxa Abundance DFs

```{r taxaAbundance}
#----- Make Ground Truth Plots for Each Taxa of Interest -----#

taxaAbundance <- vector(mode = "list", length = nrow(allTaxaDF))

for (i in 1:nrow(allTaxaDF)) {
  
  currentASV <- allTaxaDF[i, "ASV"]
  
  # Get name
  currentASVInfo <- allTaxaDF %>% 
    filter(ASV == currentASV) %>% 
    select(Family, Genus, Species)
  
  currentASVName <- paste(currentASVInfo, collapse = " ")
  
  currentASVAbdTable <- physeqBacteria %>% 
    psmelt() %>% 
    filter(OTU == currentASV)
  
  taxaAbundance[[i]] <- currentASVAbdTable
  names(taxaAbundance)[i] <- currentASVName
  
}
```

```{r save-taxa-abundance, eval=FALSE}
saveRDS(taxaAbundance,
        "../data/RDataObjects/16S/taxaAbundance.RDS")

dfPaths <- paste0("../analysis/results/16S/taxa_abundances/",
                  str_replace_all(names(taxaAbundance), " ", "_"),
                  "_abundances.txt")

walk2(.x = taxaAbundance,
      .y = dfPaths,
      .f = ~ write.table(x = .x, file = .y, quote = FALSE, row.names = FALSE,
                         sep = "\t"))
```

### Plot Ground Truth

```{r make-ground-truth-plots}
groundTruthPlots <- map2(.x = taxaAbundance,
                         .y = names(taxaAbundance),
                         .f = ~ PlotGroundTruth(abundanceDF = .x,
                                                varName = "timepoint_condition",
                                                taxaName = .y,
                                                subtitleStr = "All Samples"))

groundTruthPlots
```

```{r save-ground-truth-plots, include=FALSE, eval=FALSE}
gtPlotPaths <- paste0(figPath, "deseq_ground_truth_plots")

if (!file.exists(gtPlotPaths)) {
  dir.create(gtPlotPaths)
}

gtPlotNames <- paste0(gtPlotPaths, "/", 
                      str_replace_all(names(groundTruthPlots), " ", "_"),
                      "-deseq-groundTruth.pdf")

walk2(.x = groundTruthPlots,
      .y = gtPlotNames,
      .f = ~ ggsave(filename = here(.y), plot = .x, width = 11, height = 8.5,
                    units = "in"))
```

```{r labeledGT, eval=FALSE, include=FALSE}
# L. fissicatena
lfissPlot <- groundTruthPlots$`Lachnospiraceae [Eubacterium] fissicatena group NA` +
  geom_text_repel(data = subset(taxaAbundance$`Lachnospiraceae [Eubacterium] fissicatena group NA`,
                                Abundance > 1000),
                  aes_string(label = "Sample"), color = "grey30",
                  size = 4)
lfissPlot
ggsave(filename = here(paste0(gtPlotPaths, "/L_fissicatena_labeled_groundTruth.pdf")),
       plot = lfissPlot, width = 11, height = 8.5, units = "in")

# E. innoccum
einnocPlot <- groundTruthPlots$`Erysipelotrichaceae [Clostridium] innocuum group NA` +
  geom_text_repel(data = subset(taxaAbundance$`Erysipelotrichaceae [Clostridium] innocuum group NA`,
                                Abundance > 3000),
                  aes_string(label = "Sample"), color = "grey30",
                  size = 4)
einnocPlot
ggsave(filename = here(paste0(gtPlotPaths, "/E_innocuum_labeled_groundTruth.pdf")),
       plot = einnocPlot, width = 11, height = 8.5, units = "in")
```

### Run Statistics

```{r run-statistics-on-taxaAbundance, eval=FALSE}
# Run Kruskal test 
# I don't like running it on the plots because of problems with log-scaling axes
kruskalRes <- map(.x = taxaAbundance,
                  .f = ~ kruskal.test(Abundance ~ timepoint_condition,
                                      data = .x))

kruskalP <- unlist(map(.x = kruskalRes, .f = ~ .x$p.value)) %>%
  enframe(name = "taxon", value = "kruskal_p") %>% 
  mutate("is_significant" = ifelse(kruskal_p <= 0.05, yes = TRUE, no = FALSE))

# Run Dunn's post-hoc
dunnRes <- map(.x = taxaAbundance,
               .f = ~ dunnTest(x = Abundance ~ timepoint_condition,
                               data = .x,
                               method = "bonferroni"))
dunnResDF <- map(.x = dunnRes, .f = ~ .x$res)
dunnResDFModified <- map2(.x = dunnResDF,
                          .y = names(dunnResDF),
                          .f = ~ .x %>% mutate("taxon" = .y) %>% 
                            select(taxon, everything()))

```

```{r save-statistics, eval=FALSE, include=FALSE}
write_xlsx(x = kruskalP,
           path = here("../analysis/results/16S/ground_truth_p_values.xlsx"))

originalTaxaNames <- names(dunnResDFModified)
modifiedTaxa <- names(dunnResDFModified) %>% 
  str_remove_all("[^[:alnum:]]")

names(dunnResDFModified) <- modifiedTaxa

write_xlsx(x = dunnResDFModified,
           path = here("../analysis/results/16S/ground_truth_dunnPostHoc_results.xlsx"))
```

## Ground Truth Plot for Lachnospiraceae

We want to investigate the abundance of the Lachno family across the four groups.

```{r lachno-gt-plot}
lachnoAbdDF <- physeqBacteria %>% 
  tax_glom(taxrank = "Family") %>% 
  psmelt() %>% 
  filter(Family == "Lachnospiraceae")

lachnoGroundTruth <- PlotGroundTruth(abundanceDF = lachnoAbdDF,
                                     varName = "timepoint_condition",
                                     taxaName = "Family Lachnospiraceae",
                                     subtitleStr = "All Samples")
lachnoGroundTruth
```
```{r run-statistics-on-lachno-family}
lachnoKruskal <- kruskal.test(Abundance ~ timepoint_condition,
                              lachnoAbdDF)
# p-val is sig at 0.01525

lachnoDunns <- dunnTest(x = Abundance ~ timepoint_condition,
                        lachnoAbdDF, method = "bonferroni")
lachnoDunnsRes <- lachnoDunns[["res"]]
```

```{r save-lachnoGroundTruth, eval=FALSE}
ggsave(filename = here("../analysis/figures/16S/Lachnospiraceae_abundance.pdf"), 
       plot = lachnoGroundTruth, width = 11, height = 8.5, units = "in")

write_xlsx(x = lachnoDunnsRes,
           path = here("../analysis/results/16S/Lachnospiraceae_dunnPostHoc_results.xlsx"))
```


